Errorless irrationality: removing error-driven components from the inverse base-rate effect paradigm

### 2023-01-11 docs/manuscript.tex
write abstract and do small edits

\begin{abstract}
Include no author information in the initial submission, to facilitate
blind review.

@@

The \textit{inverse base-rate effect} \cite<IBRE, >{medin1988problem} is an irrational tendency in humans to overweigh rare events when faced with ambiguity.
In a traditional design, people learn to categorise two overlapping sets of features under two distinct labels.
These sets share a single feature, A, and possess a unique feature, B and C, predictive of their respective category label.
During learning, these sets of features occur at different frequencies.
The features under the common label usually occur three times as much as features under the rare label \cite{kruschke1996base}.
Following training, people categorise features presented by themselves and un unique combinations.
People optimally label uniquely predictive features, B and C, presented individually with their respective common and rare labels.
Responses on the shared feature A also show base-rate following.
But when uniquely predictive features are paired, B and C, people tend to respond with the rare category label.
According to Classical Probability Theory, the rational response is to attribute the common label to this ambiguous compound, because it is the most frequently occurring label.
This rare bias on ambiguous combinations of BC has been observed across a different variety of experimental manipulations \cite{kalish2001inverse,don2017effects,don2021attention,inkster2022effect,wills2014attention}.
For a more thorough introduction to this irrational bias, see a review by \citeA{don2021hearing}.

@@

The most prominent theories of the inverse base-rate effect involve an attentional mechanism that drives not only learning but responding as well.
These theories are formal models.
They are: a neural network with an exemplar-mediated attention to distinctive input, EXIT \cite{kruschke2001toward}, a three-layer neural network with competitive attentional gating, and a four-layer neural network with an additional rapid attentional shift \cite{paskewitz2020dissecting}.
All these explanations rely on a process that reallocates attention in response to prediction errors - they update attentional values according to gradient descent.
Their explanation is simple.
During learning, people learn to label the AB compound first.
They are still learning to label the AC compound, so when they make an error, attention relocates towards the uniquely predictive feature C to reduce future errors.
This results in C acquiring higher attentional salience than B.
When the ambiguous BC compound is presented, C will dominate responding, resulting in an irrational tendency to respond with the rare label.
According to these models, this irrationality results from an optimisation process that tries to reduce the errors people make.
This process results in an asymmetric representation that can be summarised as AB belongs to common, $AB \to common$, and C belongs to rare, $C \to rare$ \cite{kruschke2001inverse}.

@@

To our knowledge, there is only one attempt to implement the IBRE procedure without explicit feedback.
In their Experiment 2, \citeA{johansen2007paradoxical} tried to observe the inverse base-rate effect both in a predictive-learning condition and in an observational learning condition.
In the observational learning condition, category labels and features were presented at the same time to participants.
Their design involved disjoint cues (where categories shared no features in common), while the canonical design depends on a shared feature during training.
According to the theories of the IBRE, this shared feature facilitates attentional relocation.
This attentional tuning in turn pushes responding toward the rare label.
In contrast, \citeA{johansen2007paradoxical} trained participants on $AB \to common$ and $C \to rare$.
% LD: maybe include here some info on the shared-cue effect?
Their design was optimized to investigate the hypothesised asymmetric cognitive representation of the two categories.
As a result, the only instance when they observed a rare bias was when the common feature, $B$ ,presented in compound during training was paired with a rare, $C$, feature presented by itself during training.
This provided evidence for the asymmetric cognitive representation \cite{kruschke2001inverse} hypothesised to develop during training.
% LD: this speculation might be redundant?
A simple neural-network explanation can account for this by positing that the rare feature developed stronger connections with the category label.
Compound features share the connection with the category label, so any update to these connections dissipates between them.
There is no need to relocate attention to reduce errors, therefore attention will not bias responding toward the rare label.
But stronger $C \to rare$ feature-label connections will result in a rare bias on BC compounds during the test phase.
In order to investigate the role of prediction error in response to feedback, we need to tweak their observational learning condition to conform to a more canonical implementation of the procedure.
This will involve reimplementing the shared feature.
In terms of a clear observational-learning version of the IBRE, \citeA{johansen2007paradoxical} included the result of a short pilot experiment in their Appendix.
There are no details about the procedure of this experiment, so we cannot make direct comparisons.
We will follow up on that line of investigation.

@@

In both experiments, the IBRE occured without any explicit detail in the experimental procedure that would result in prediction error.
Therefore, any theorised error-driven process must be able to operate without explicit feedback.
Most prominent theories and their corresponding formal specification rely on relocating attention in response to prediction error.
They are unable to process the current experiment, because they are not designed to encode information presented without feedback or fit cued recall tasks.

The two experiments suggest that the necessary conditions to observe the IBRE are fewer than previously established.
In Experiment 2, the only remaining conditions are the two uniquely predictive features, an overlapping feature, sequential presentation and the base rate.
\cite{johansen2007paradoxical} gave evidence that even the overlapping features are not necessary.
In their design, common category exemplars were made up of two features, while rare exemplars only had a single feature - disjoint cues.
This can mean that common exemplars need to be composed of at least two features, while the rare exemplar must have only one feature.
If the rare exemplar has one uniquely predictive feature and another feature that is not shared with the other category, we do not observe the effect.
This is the shared-cue effect \cite{kruschke2001inverse, wills2014attention}.
One can hypothesise that the shared cue and the disjoint cue are sufficient conditions, but at least one of them is necessary.
In addition, one can hypothesise that in the absence of a shared feature, common exemplars must contain more features than rare exemplars.

% LD: so it seems that Johansen et al treated asymmetric representation as an experimental condition and not as a theoretical construct
% LD: I think this is a problem because asymmetric representation is hypothesized to be an asymmetry in the way the two categories represented
Compared to \citeA{johansen2007paradoxical}, we do not consider the asymmetric representation and the base-rate neglect as necessary conditions.
From our point of view, they are theoretical constructs.
They are not part of the experimental design.
We consider them to result from the environmental conditions specified in the experiment required for the IBRE as opposed to being necessary/sufficient conditions for the IBRE.

% Asymmetric representation can arise in different ways.
One hypothesised way asymmetric representation is manifested is the attentional tuning of cognitive representation of category exemplars.
This is not necessarily absent in our experiment, but is not directly tested.
Our experiments do not give direct evidence against the role of attention in developing the asymmetric representation.
Nonetheless, it must not happen through an error-driven process as conceptualised in most prominent theories.
To further investigate this, the cued-recall IBRE procedure could incorporate eye-tracking to measure dwell time and order of information encoding.
Even if we observe more and longer fixations to C relative to B, attention can still be allocated asymmetrically to drive the distinguishability of categories.
Explanations do not need to invoke an explicit error-driven process.

- auxiliary phenomenon (Wills CIRP addition)
- Things that any theory of IBRE should explain
- current theories fall short
- eye-tracking and attention
-

@@

In Experiment 1, we conducted a successful conceputal replication of \citeA{johansen2007paradoxical}, which gave evidence for the IBRE being independent of supervised learning procedures.
In addition, Experiment 2 further suggests that the IBRE generalises beyond simple predictive-learning \cite{kruschke1996base,don2021hearing} and decision-making \cite{johansen2007paradoxical} paradigms.
Theories of IBRE are inadequate to account for these findings, because of their inability to extend beyond supervised learning.

>>>>

\begin{abstract}

The inverse base-rate effect is a robust irrational bias that arises when people face ambiguity.
The most prominent theories of this irrational bias depend on prediction error.
In this study, we gradually removed elements of a predictive learning design.
In our first experiment, we removed explicit feedback by implementing the inverse base-rate effect in an observational learning procedure. 
In our second study, we further removed any causal relationship between stimulus features and category labels.
This removed any information participants could use to identify category labels.
In both experiments, the inverse base-rate effect persisted and remained robust.
This outcome suggests that this irrational bias is independent of supervised learning procedures.
We propose that any theories and models of the inverse base-rate effect must be designed in a way that allows information encoding and connection updates to proceed without prediction error.
We end by proposing two clear paths for future investigations.

@@

The \textit{inverse base-rate effect} \cite<IBRE, >{medin1988problem} is an irrational tendency in humans to overweigh rare events when faced with ambiguity.
In a traditional design, people learn to categorise two overlapping sets of features under two distinct labels.
These sets share a single feature, A, and possess a unique feature, B and C, predictive of their respective category label.
During learning, these sets of features occur at different frequencies.
The features under the common label usually occur three times as much as features under the rare label \cite{kruschke1996base}.
Following training, people categorise features presented by themselves and in novel combinations.
People tend to optimally label uniquely predictive features, $B$ and $C$, with their respective common and rare labels when presented by themselves.
Responses on the shared feature A tend to show base-rate following.
But when uniquely predictive features are paired, $BC$, people tend to respond with the rare category label.
According to Classical Probability Theory, the rational response is to attribute the common label to this ambiguous compound, because it is the most frequently occurring label.
This rare bias on ambiguous combinations of BC has been observed across a different variety of experimental manipulations \cite{kalish2001inverse,don2017effects,don2021attention,inkster2022effect,wills2014attention}.
For a more thorough introduction to this irrational bias, see a review by \citeA{don2021hearing}.

@@

% LD: if I talk about supervised/predictive learning at the end, I should probably include something here
The most prominent theories of the inverse base-rate effect involve an attentional mechanism that drives not only learning but responding as well.
These theories are formal models.
They are: a neural network with an exemplar-mediated attention to distinctive input, EXIT \cite{kruschke2001toward}, a three-layer neural network with competitive attentional gating, and a four-layer neural network with an additional rapid attentional shift \cite{paskewitz2020dissecting}.
All these explanations rely on a process that relocates attention in response to prediction errors - they update attentional values according to gradient descent.
Their explanation is simple.
During learning, people learn to label the AB compound first.
They are still learning to label the AC compound, so when they make an error, attention relocates towards the uniquely predictive feature C to reduce future errors.
This results in C acquiring higher attentional salience than B.
When the ambiguous BC compound is presented, C will dominate responding, resulting in an irrational tendency to respond with the rare label.
According to these models, this irrationality results from an optimisation process that tries to reduce the errors people make.
This process results in an asymmetric representation that can be summarised as AB belongs to common, $AB \to common$, and C belongs to rare, $C \to rare$ \cite{kruschke2001inverse}.

@@

To our knowledge, there is only one attempt to implement the IBRE procedure without explicit feedback.
In their Experiment 2, \citeA{johansen2007paradoxical} tried to observe the inverse base-rate effect both in a predictive-learning condition and in an observational learning condition.
In the observational learning condition, category labels and features appeared together at the same time on the computer screen.
Their design involved disjoint cues (where categories shared no features in common), while the canonical design depends on a shared feature during training \cite{kruschke2001inverse, wills2014attention}.
According to the theories of the IBRE, this shared feature facilitates attentional relocation.
This attentional tuning in turn pushes responding toward the rare label.
In contrast, \citeA{johansen2007paradoxical} trained participants on $AB \to common$ and $C \to rare$.
% LD: maybe include here some info on the shared-cue effect?
Their design was optimized to investigate the hypothesised asymmetric cognitive representation of the two categories.
As a result, the only instance when they observed a rare bias was when the common feature, $B$ ,presented in compound during training was paired with a rare, $C$, feature presented by itself during training.
This provided evidence for the asymmetric cognitive representation \cite{kruschke2001inverse} hypothesised to develop during training.
% LD: this speculation might be redundant?
A simple neural-network explanation can account for this by positing that the rare feature developed stronger connections with the category label.
Compound features share the connection with the category label, so any update to these connections dissipates between them.
There is no need to relocate attention to reduce errors, therefore attention will not bias responding toward the rare label.
But stronger $C \to rare$ feature-label connections will result in a rare bias on BC compounds during the test phase.
In order to investigate the role of prediction error in response to feedback, we need to tweak their observational learning condition to conform to a more canonical implementation of the procedure.
This will involve reimplementing the shared feature.
In terms of a clear observational-learning version of the IBRE, \citeA{johansen2007paradoxical} included the result of a short pilot experiment in their Appendix.
There are no details about the procedure of this experiment, so we cannot make direct comparisons.
We will follow up on that line of investigation.

@@

In both experiments, the IBRE occurred without any explicit detail in the experimental procedure that would result in prediction error.
Therefore, any theorised error-driven process must be able to operate without explicit feedback.
Most prominent theories and their corresponding formal specification rely on relocating attention in response to prediction error.
They are unable to process the current experiment because they are not designed to encode information presented without feedback or fit cued recall tasks.

The two experiments suggest that the necessary conditions to observe the IBRE are fewer than previously established.
% LD: so it seems that Johansen et al treated asymmetric representation as an experimental condition and not as a theoretical construct
% LD: I think this is a problem because asymmetric representation is hypothesized to be an asymmetry in the way the two categories represented
Compared to \citeA{johansen2007paradoxical}, we also do not consider asymmetric representation and base-rate neglect as necessary conditions.
We consider them to result from the environmental conditions specified in the experiment required for the IBRE as opposed to being necessary/sufficient conditions for the IBRE.
From our point of view, they are theoretical constructs.
They are not part of the experimental design.

In Experiment 2, the only remaining conditions are the two uniquely predictive features, an overlapping feature, sequential presentation and the base rate.
\cite{johansen2007paradoxical} gave evidence that even the overlapping features are not necessary.
In their design, common category exemplars were made up of two features, while rare exemplars only had a single feature - disjoint cues.
This can mean that common exemplars need to be composed of at least two features, while the rare exemplar must have only one feature.
If the rare exemplar has one uniquely predictive feature and another feature that is not shared with the other category, we do not observe the effect.
This is the shared-cue effect \cite{kruschke2001inverse, wills2014attention}.
One can hypothesise that the shared cue and the disjoint cue are sufficient conditions, but at least one of them is necessary.
In addition, one can hypothesise that in the absence of a shared feature, common exemplars must contain more features than rare exemplars.

% Asymmetric representation can arise in different ways.
One hypothesised way asymmetric representation is manifested is the attentional tuning of cognitive representation of category exemplars.
This is not necessarily absent in our experiments but is not directly tested.
Our experiments do not give direct evidence against the role of attention in developing asymmetric representation.
Nonetheless, it must not happen through an error-driven process as conceptualised in most prominent theories.
To further investigate this, the cued-recall IBRE procedure could incorporate eye-tracking to measure dwell time and order of information encoding.
Even if we observe more and longer fixations to C relative to B, attention can still be allocated asymmetrically to drive the distinguishability of categories.
Explanations do not need to invoke an explicit error-driven process.

@@

Across two experiments, we investigated whether the IBRE requires error-driven components of the experimental procedure.
In Experiment 1, we conducted a successful conceputal replication of \citeA{johansen2007paradoxical}, which gave evidence for the IBRE being independent of supervised learning procedures.
In addition, Experiment 2 further suggests that the IBRE generalises beyond simple predictive-learning \cite<e.g. >{medin1988problem,kruschke1996base,wills2014attention} and decision-making \cite{johansen2007paradoxical} paradigms.
This further suggest prediction error in terms of explicit feedback is not a necessary condition.
Theories of IBRE are inadequate to account for these findings, because of their inability to extend beyond supervised learning.
